---
title: "brain_expression_temporal_dynamics"
output: html_document
date: '2022-05-06'
---

# **Temporal changes of gene expression in health and mental disorders**

#### Abstract:
Schizophrenia (SCZ), bipolar disorder (BD), and major depressive disorder (MDD), are complex and heterogeneous conditions thought to arise from an interplay of diverse genetic and environmental factors. The molecular events contributing to disease development, manifestation, and course are known to be distributed through embryonic life to the elderly. However, due to the age of onset, clinical samples are always thresholded to relatively older age, and little is known about early dynamics in gene expression in these diseases. 
We performed a secondary analysis of microarray datasets of post-mortem prefrontal cortex of control brains and patients with SCZ, BD, and MDD available in Gene Expression Omnibus and the Stanley Medical Research Institute Online Genomics Database․ We used self-organized maps machine learning (oposSOM package) to dissect brain transcriptome into functional gene modules associated with aging and diseases and Gaussian Process Regression (DE_time package) to detect time-perturbation points of disease-associated functional points characteristics to mental disorders. Finally, we performed eQTL enrichment of genes in time-perturbed functional modules. Our results indicate that mental disorders are characterized by early, mid, and late deregulation of identified functional spots. The results align with the hypothesis that two or more 'hits' are required over the lifespan rather than only one early-life event for disease manifestation. Time-perturbed functional spots were enriched with eQTL genes implicating the contribution of genetic components in the gene expression dynamics and development of the disease phenotype.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning = FALSE)  #eliminate code, warnings and messages in the knitted markdown, keep all results of chunks (text and plots)
```

**Loading needed libraries**

```{r}
library(sva)
library(RColorBrewer)
library(dplyr)
library(Matrix)
library(AtmRay) 
library(spam)
library(gptk)
library(DEtime) 
library(enrichR)
library(limma)
library(venn)
library(networkD3)
library(DOSE)
library(org.Hs.eg.db)
library(here)
library(inborutils)
library(utils)
library(oposSOM)
```

```{r}
##initiate the working directory to be the dir of the running file
here()
```

## **Downloading and Extracting Data from Zenodo**

```{r}
##skip if data has been downloaded
download_zenodo('10.5281/zenodo.7936218') #takes around 2 hours
unzip("./Data.zip", exdir = "./Data")
```

**Function to read pheno data**

```{r}
load_pheno <-  function(where = "Data/pheno/", batch) {
    ##reading pheno data
  phenos.full <- dir(where, pattern = "pheno.csv", full.names = T)
  phenos <- dir(where, pattern = "pheno.csv")
  pheno.names <- sapply(phenos, function(x) strsplit(x,split = ".", fixed = T )[[1]][1])
  names(pheno.names) <- NULL
  
  pheno.list <- list()
  for(i in 1:length(phenos.full)) {
    pheno.list[[i]] <- read.csv(file <- phenos.full[[i]], header = T, row.names = 1)
  }
  
  
  age <- vector(mode = "numeric", length = length(batch))
  sex <- vector(mode = "character", length = length(batch))
  ph  <- vector(mode = "numeric", length = length(batch))
  pmi <- vector(mode = "numeric", length = length(batch))
  
  
  names(age) <- names(sex) <- names(ph) <- names(pmi) <- names(batch)
  
  for(i in 1:length(pheno.list)){
    samples.tmp <- intersect(rownames(pheno.list[[i]]), names(batch))
    age[samples.tmp] <- pheno.list[[i]][samples.tmp, "age"]
    sex[samples.tmp] <- pheno.list[[i]][samples.tmp, "sex"]  
    ph[samples.tmp] <- as.numeric(pheno.list[[i]][samples.tmp, "ph"])
    pmi[samples.tmp] <- as.numeric(pheno.list[[i]][samples.tmp, "pmi"])
  }
  
  return(list(age, sex, ph, pmi))
}
```

# **PCA and Batch Effects Analysis** 

-   Loading data and removing batch effects of covariates

```{r}
load("Data/Brain_development_v3.Rdata")

dat <- do.call(cbind, indata.list)
batch <- rep(1:length(group.labels.list), times = sapply(group.labels.list, length))

group.labels <- do.call(c, group.labels.list) 
group.label.names <- sapply(names(group.labels), function(x) strsplit(x, split = ".", fixed = T)[[1]][2])
names(group.label.names) <- NULL
names(group.labels) <- group.label.names
names(batch) <- names(group.labels)

##loading phenotype data
pheno_vars <- load_pheno(where="Data/pheno",batch)
age <- pheno_vars[[1]]
sex <- pheno_vars[[2]]
ph <- pheno_vars[[1]]
pmi <- pheno_vars[[1]]

pmi[is.na(pmi)] <- mean(pmi, na.rm = T)
ph[is.na(ph)] <- mean(ph, na.rm = T)

dat <- dat - rowMeans(dat)
mod <- model.matrix(~0+as.factor(group.labels))

##removing batch effects
datn <- removeBatchEffect(dat, batch = batch, batch2 = sex, covariates = cbind(ph, pmi), design = mod)

col.pal.batch <- RColorBrewer::brewer.pal(n = length(unique(batch)), name = "Accent")
c.ind <- grep("Cntrl", group.labels) 
s.ind <- grep("Scz", group.labels) 
b.ind <- grep("BP", group.labels) 
m.ind <- grep("MDD", group.labels)

condition <- rep(1, times=length(group.labels))
condition[s.ind] <- 2
condition[b.ind] <- 3
condition[m.ind] <- 4

col.pal.cond <- RColorBrewer::brewer.pal(n = length(unique(condition)), name = "Dark2")
col.batch <- col.pal.batch[batch]
col.cond <- col.pal.cond[condition]


p <- try(prcomp(dat), silent = T)
pn <- try(prcomp(datn), silent = T)

if(class(p)!="try-error") {
  {
    plot(p$rotation, col = col.batch, main = "before batch adjustment by batch")
    legend("topright", legend=unique(batch),
           col=col.pal.batch, pch = 1, cex=0.8)
  }
}

if(class(p)!="try-error") {
  {
    plot(p$rotation, col = col.cond, main = "before batch adjustment by condition")
    legend("topright", legend=unique(condition),
           col=col.pal.cond, pch = 1, cex=0.8)
  }
}

if(class(pn)!="try-error") {
  {
    plot(pn$rotation, col = col.batch, main = "after batch adjustment by batch")
    legend("topright", legend=unique(batch),
           col=col.pal.batch, pch = 1, cex=0.8)
  }
}

if(class(pn)!="try-error") {
  {
    plot(pn$rotation, col = col.cond, main = "after batch adjustment by condition")
    legend("topright", legend=unique(condition),
           col=col.pal.cond, pch = 1, cex=0.8)
  }
}

```

-   PCA for Control samples

```{r}
cntrl.ind <- grep("Cntrl", group.labels)
cntrl.dat <- dat[, cntrl.ind]
cntrl.datn <- datn[, cntrl.ind]
cntrl.batch <- batch[cntrl.ind]
delta <- min(cntrl.batch)-1
cntrl.batch.n <- cntrl.batch-delta 
cntrl.lbl <- group.labels[cntrl.ind]

ord <- unique(cntrl.lbl)[c(1,2,3,4,5,6,8,7,9)]
ind <- NULL
for(i in ord){
  ind <- c(ind, grep(pattern = i, cntrl.lbl, fixed = T) )
}

cntrl.ind <- cntrl.ind[ind]
cntrl.dat <- cntrl.dat[, ind]
cntrl.datn <- cntrl.datn[, ind]
cntrl.batch <- cntrl.batch.n[ind]
cntrl.lbl <- cntrl.lbl[ind] 

# col.pal <- RColorBrewer::brewer.pal(n = length(unique(cntrl.batch.n)), name = "Accent")
# cols <- col.pal[cntrl.batch.n]
# 
# p <- try(prcomp(cntrl.dat-rowMeans(cntrl.dat)), silent = T)
# if(class(p)!="try-error") {
#   {
#     plot(p$rotation, col = cols, main = "before batch adjustment")
#     legend("topright", legend=unique(cntrl.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }
# 
# pn <- try(prcomp(cntrl.datn), silent = T)
# if(class(pn)!="try-error") {
#   {
#     plot(pn$rotation, col = cols, main = "after batch adjustment")
#     legend("topright", legend=unique(cntrl.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }


```

-   PCA for Scz samples

```{r}

scz.ind <- grep("Scz", group.labels)
scz.dat <- dat[, scz.ind]
scz.datn <- datn[, scz.ind]
scz.batch <- batch[scz.ind]
delta <- min(scz.batch)-1
scz.batch.n <- scz.batch-delta 
scz.lbl <- group.labels[scz.ind]

o <- order(scz.lbl)
scz.datn <- scz.datn[, o]
scz.batch <- scz.batch[o]
scz.lbl <- scz.lbl[o] 
# col.pal <- RColorBrewer::brewer.pal(n = length(unique(scz.batch.n)), name = "Accent")
# cols <- col.pal[scz.batch.n]
# 
# 
# p <- try(prcomp(scz.dat-rowMeans(scz.dat)), silent = T)
# if(class(p)!="try-error") {
#   {
#     plot(p$rotation, col = cols, main = "before batch adjustment")
#     legend("topright", legend=unique(scz.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }
# 
# pn <- try(prcomp(scz.datn), silent = T)
# if(class(pn)!="try-error") {
#   {
#     plot(pn$rotation, col = cols, main = "after batch adjustment")
#     legend("topright", legend=unique(scz.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }
```

-   PCA for BP samples

```{r}
bp.ind <- grep("BP", group.labels)
bp.dat <- dat[, bp.ind]
bp.datn <- datn[, bp.ind]
bp.batch <- batch[bp.ind]
delta <- min(bp.batch)-1
bp.batch.n <- bp.batch-delta 
bp.lbl <- group.labels[bp.ind]

o <- order(bp.lbl)
bp.datn <- bp.datn[, o]
bp.batch <- bp.batch[o]
bp.lbl <- bp.lbl[o] # col.pal <- RColorBrewer::brewer.pal(n = length(unique(bp.batch.n)), name = "Accent")
# cols <- col.pal[bp.batch.n]
# 
# 
# p <- try(prcomp(bp.dat-rowMeans(bp.dat)), silent = T)
# if(class(p)!="try-error") {
#   {
#     plot(p$rotation, col = cols, main = "before batch adjustment")
#     legend("topright", legend=unique(bp.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }
# 
# pn <- try(prcomp(bp.datn), silent = T)
# if(class(pn)!="try-error") {
#   {
#     plot(pn$rotation, col = cols, main = "after batch adjustment")
#     legend("topright", legend=unique(bp.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }
```

-   PCA for MDD samples

```{r}
mdd.ind <- grep("MDD", group.labels)
mdd.dat <- dat[, mdd.ind]
mdd.datn <- datn[, mdd.ind]
mdd.batch <- batch[mdd.ind]
delta <- min(mdd.batch)-1
mdd.batch.n <- mdd.batch-delta 
mdd.lbl <- group.labels[mdd.ind]

o <- order(mdd.lbl)
mdd.datn <- mdd.datn[, o]
mdd.batch <- mdd.batch[o]
mdd.lbl <- mdd.lbl[o]
# col.pal <- RColorBrewer::brewer.pal(n = length(unique(mdd.batch.n)), name = "Accent")
# cols <- col.pal[mdd.batch.n]
# 
# 
# p <- try(prcomp(mdd.dat-rowMeans(mdd.dat)), silent = T)
# if(class(p)!="try-error") {
#   {
#     plot(p$rotation, col = cols, main = "before batch adjustment")
#     legend("topright", legend=unique(mdd.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }
# 
# pn <- try(prcomp(mdd.datn), silent = T)
# if(class(pn)!="try-error") {
#   {
#     plot(pn$rotation, col = cols, main = "after batch adjustment")
#     legend("topright", legend=unique(mdd.batch.n),
#            col=col.pal, pch = 1, cex=0.8)
#   }
# }
```

-   Combine data and save

```{r}
indata <- cbind(cntrl.datn, scz.datn, bp.datn, mdd.datn)
group.labels <- c(cntrl.lbl, scz.lbl, bp.lbl, mdd.lbl)
batch <- c(cntrl.batch, scz.batch, bp.batch, mdd.batch)
# save(list = c("indata", "group.labels", "batch"), file = "som_input_v5.Rdata")
```

# **SOM training and results**

```{r}
## run this chunk if need to regenerate SOM results from preprocessed data
##source("script_SOM_v5.R")
```

**Loading SOM results and combining spots**

```{r, include=FALSE}
##loading SOM result
load("Data/v5_results.SOM - Results/v5_results.SOM.RData")

# combined spots
env$spot.list.GSZ.scores$spotdata <- env$samples.GSZ.scores

km.spot <- c("W", "S", "G", "U", "B", "I")
ov.spot <- c("U", "Q", "G", "A")
gr.spot <- c("K", "I")
dm.spot <- c("V1", "J1", "E1", "J", "P", "S")

names_km.spot <- paste0("KM_", c("W", "S", "G", "U", "B", "I"))
names_ov.spot <- paste0("OV_", c("U", "Q", "G", "A"))
names_gr.spot <- paste0("GR_", c("K", "I"))
names_dm.spot <- paste0("DM_", c("V1", "J1", "E1", "J", "P", "S"))

spot.list <- list()
spot.list$spots <- c(env$spot.list.kmeans$spots[km.spot],
                     env$spot.list.overexpression$spots[ov.spot],
                     env$spot.list.group.overexpression$spots[gr.spot],
                     env$spot.list.dmap$spots[dm.spot])
names(spot.list$spots) <- c(names_km.spot, names_ov.spot, names_gr.spot, names_dm.spot)

spot.list$spotdata <- rbind(env$spot.list.kmeans$spotdata[km.spot, ],
                            env$spot.list.overexpression$spotdata[ov.spot, ],
                            env$spot.list.group.overexpression$spotdata[gr.spot, ],
                            env$spot.list.dmap$spotdata[dm.spot, ])

rownames(spot.list$spotdata) <-  c(names_km.spot, names_ov.spot, names_gr.spot, names_dm.spot)

spot.list$overview.map <-  apply(apply(env$metadata, 2, function(x) { (x - min(x)) / (max(x) - min(x)) }), 1, max)

overview.mask <- rep(NA, times = env$preferences$dim.1stLvlSom^2)

for(i in 1:length(rownames(spot.list$spotdata)) ) {
  
  spot.type <- rownames(spot.list$spotdata)[i]
  spot.type <- strsplit(spot.type, split = "_", fixed = TRUE)
  
  if(spot.type[[1]][1]=="KM") {
    ind.orig.spot <- which(names(env$spot.list.kmeans$spots)==spot.type[[1]][2])
    ind.map <- which(env$spot.list.kmeans$overview.mask==ind.orig.spot)
    overview.mask[ind.map] <- i
  } else if (spot.type[[1]][1]=="OV") {
    ind.orig.spot <- which(names(env$spot.list.overexpression$spots)==spot.type[[1]][2])
    ind.map <- which(env$spot.list.overexpression$overview.mask==ind.orig.spot)
    overview.mask[ind.map] <- i
  } else if (spot.type[[1]][1]=="GR") {
    ind.orig.spot <- which(names(env$spot.list.group.overexpression$spots)==spot.type[[1]][2])
    ind.map <- which(env$spot.list.group.overexpression$overview.mask==ind.orig.spot)
    overview.mask[ind.map] <- i
  } else if (spot.type[[1]][1]=="DM") {
    ind.orig.spot <- which(names(env$spot.list.dmap$spots)==spot.type[[1]][2])
    ind.map <- which(env$spot.list.dmap$overview.mask==ind.orig.spot)
    overview.mask[ind.map] <- i
  }
} 

spot.list$overview.mask <- overview.mask

names(spot.list$spots) <- LETTERS[1:length(names(spot.list$spots))]
rownames(spot.list$spotdata) <- LETTERS[1:length(names(spot.list$spots))]


dir.name <- "combined.spots"

if(!dir.exists(dir.name)) {
  dir.create(dir.name)
}
```

# **Differentially expressed genes across age groups in studied diseases**

```{r}

gene.files <- dir("Data/v5_results.SOM - Results/Summary Sheets - Differences/CSV Sheets/Gene Lists - Global/", full.names = T)
gene.files.names <- dir("Data/v5_results.SOM - Results/Summary Sheets - Differences/CSV Sheets/Gene Lists - Global/")
gene.files.names <- sapply(gene.files.names, function(x) strsplit(x, split=".", fixed=T)[[1]][1])
names(gene.files.names) <- NULL
diff.gene.list <- vector(mode = "list", length = length(gene.files.names))
for(i in 1:length(diff.gene.list)){
  dat <- read.csv(file=gene.files[i],
                  header = T, skip=16)
  fdr <- sapply(dat$fdr, function(x) strsplit(x, split = " ", fixed = T)[[1]][1])
  names(fdr) <- NULL
  fdr <- as.numeric(fdr)
  diff.gene.list[[i]] <- list(up=unique(dat$Symbol[dat$logFC>0 & fdr < 0.05]),
                              down=unique(dat$Symbol[dat$logFC<0 & fdr < 0.05]) )
}
names(diff.gene.list) <- gene.files.names


gs.files <- dir("Data/v5_results.SOM - Results/Summary Sheets - Differences/CSV Sheets/Gene Set Lists - Global/", full.names = T)
gs.files.names <- dir("Data/v5_results.SOM - Results/Summary Sheets - Differences/CSV Sheets/Gene Set Lists - Global/")
gs.files.names <- sapply(gs.files.names, function(x) strsplit(x, split=".", fixed=T)[[1]][1])
names(gs.files.names) <- NULL

diff.gs.list <- vector(mode = "list", length = length(gene.files.names))
for(i in 1:length(diff.gs.list)){
  dat <- read.csv(file=gs.files[i],
                  header = T)
  fdr.u <- sapply(dat$fdr, function(x) strsplit(x, split = " ", fixed = T)[[1]][1])
  names(fdr.u) <- NULL
  fdr.u <- as.numeric(fdr.u)
  fdr.d <- sapply(dat$fdr., function(x) strsplit(x, split = " ", fixed = T)[[1]][1])
  names(fdr.d) <- NULL
  fdr.d <- as.numeric(fdr.d)
  diff.gs.list[[i]] <- list(up=dat$Upregulated[dat$GSZ > 3 & fdr.u<0.05], down=dat$Downregulated[dat$GSZ. < -3 & fdr.d<0.05])
}

names(diff.gs.list) <- gs.files.names
gs.up <- sapply(diff.gs.list, function(x) length(x$up))
gs.down <- sapply(diff.gs.list, function(x) length(x$down))
genes.up <- sapply(diff.gene.list, function(x) length(x$up))
genes.down <- sapply(diff.gene.list, function(x) length(x$down))
par(mar=c(10,10,5,5))
genes.mat <- rbind(genes.up, genes.down)
barplot(genes.mat[,1:4], col=c("red", "blue"), las=3)
barplot(genes.mat[,5:7], col=c("red", "blue"), las=3)
barplot(genes.mat[,8:12], col=c("red", "blue"),las=3)
gs.mat <- rbind(gs.up, gs.down)
barplot(gs.mat[,1:4], col=c("red", "blue"), las=3)
barplot(gs.mat[,5:7], col=c("red", "blue"), las=3)
barplot(gs.mat[,8:12], col=c("red", "blue"), las=3)

```


# **scRNA-seq data and spot correlation**

### **scRNA cell marker gene signatures and spot correlation**

```{r}
#scRNA-seq signatures
load("Data/scRNA-Ruzicka_data/scRNA_cell_markers.Rdata")

scRNA.list = c(scRNA_cell_markers)

# scRNA.list = q$KM_G
pdf(file = paste0(dir.name, "/", "scRNA.cell.mapping.pdf"))
for(i in 1:length(scRNA.list)) {
  image(matrix(log10(apply(env$metadata, 1, var)), env$preferences$dim.1stLvlSom,
               env$preferences$dim.1stLvlSom), axes = FALSE, col = env$color.palette.heatmaps(1000),
        main = names(scRNA.list)[i], cex.main = 1.5)
  
  set.genes <- unique(names(env$gene.info$names)[env$gene.info$names %in% scRNA.list[[i]] ])
  show(length(set.genes))
  n.map <- matrix(0, env$preferences$dim.1stLvlSom, env$preferences$dim.1stLvlSom)
  # n.map <- rep(NA, times=env$preferences$dim.1stLvlSom*env$preferences$dim.1stLvlSom)
  gs.nodes <- env$som.result$feature.BMU[set.genes]
  n.map[as.numeric(names(table(gs.nodes)))] <- table(gs.nodes)
  n.map[which(n.map == 0)] <- NA
  n.map <- matrix(n.map, env$preferences$dim.1stLvlSom)
  lim <- c(1, env$preferences$dim.1stLvlSom) + env$preferences$dim.1stLvlSom *
    0.01 * c(-1, 1)
  colr <- env$color.palette.heatmaps(1000)[(na.omit(as.vector(n.map)) -
                                              min(n.map, na.rm = TRUE))/max(1, (max(n.map, na.rm = TRUE) - min(n.map, na.rm = TRUE))) * 999 + 1]
  par(new=T)
  plot(which(!is.na(n.map), arr.ind = TRUE), xlim = lim,
       ylim = lim, pch = 16, axes = FALSE, xlab = "", ylab = "",
       xaxs = "i", yaxs = "i", col = colr, cex = 0.5 +
         na.omit(as.vector(n.map))/max(n.map, na.rm = TRUE) *
         2.8)
}

dev.off() 
```

```{r}
scRNA.gs.def.list <- vector(mode="list", length=length(scRNA.list))

for(i in 1:length(scRNA.gs.def.list)) {
  ind <- env$gene.info$names %in% scRNA.list[[i]]
  e.genes <- unique(env$gene.info$ids[ind])
  scRNA.gs.def.list[[i]]$Genes <- e.genes
  scRNA.gs.def.list[[i]]$Type <- "scRNA"
}

names(scRNA.gs.def.list) <- names(scRNA.list)

mean.ex.all <- colMeans( env$indata.ensID.m )
sd.ex.all <- apply( env$indata.ensID.m, 2, sd )

scRNA.GSZ.scores <- t( sapply( scRNA.gs.def.list, Sample.GSZ, env$indata.ensID.m, mean.ex.all, sd.ex.all ) )

pdf(file = paste0(dir.name, "/", "scRNA.cell.boxplot.pdf")) 
par(mar=c(10,5,5,5))
for(i in 1:nrow(scRNA.GSZ.scores))
  boxplot(scRNA.GSZ.scores[i, ] ~ factor(env$group.labels, levels = unique(env$group.labels)), 
          col = env$groupwise.group.colors, main = rownames(scRNA.GSZ.scores)[i], border = T, 
          las=3, xlab = NULL)

dev.off()
```

```{r}

ind.cor <- list(ind.c = grep("Cntrl_", env$group.labels), 
                ind.s = grep("Scz_", env$group.labels), 
                ind.b = grep("BP_", env$group.labels),
                ind.m = grep("MDD_", env$group.labels))
group=c("Cntrl", "Scz", "BP", "MDD")

links <- data.frame(matrix(0,ncol = 4, nrow = 0))
colnames(links) <- c("source", "target", "group", "value")

for(i in 1:length(ind.cor)){
  x <- rbind(scRNA.GSZ.scores[, ind.cor[[i]]], spot.list$spotdata[,ind.cor[[i]]])
  y <- cor(t(x))
  y <- y[rownames(spot.list$spotdata), rownames(scRNA.GSZ.scores)]
  
  ind <- which(y>0.7, arr.ind = TRUE)
  spot_cell <- apply(ind, 1, function(x) colnames(y)[x[2]])
  spot_cell.cor <- apply(ind,1,function(x) y[x[1], x[2]])
  links.tmp <- data.frame(source=names(spot_cell), 
                          target=spot_cell,
                          group=rep(group[i], times=nrow(ind)),
                          value=round(1*spot_cell.cor, 2))
  links <- rbind(links, links.tmp)
}

nodes.df <- data.frame(names=c(unique(links$source), unique(links$target)))
nodes.df$group <- as.factor(c("node_group"))

my_color <- 'd3.scaleOrdinal() .domain(["Cntrl", "Scz", "BP", "MDD","node_group"]) .range(["#EFF3FF", "#BDD7E7", "#6BAED6", "#2171B5", "grey"])'

links$IDsource <- match(links$source, nodes.df$names)-1
links$IDtarget <- match(links$target, nodes.df$names)-1

p <- sankeyNetwork(Links = links, Nodes = nodes.df, 
                   Source = "IDsource", Target = "IDtarget", sinksRight = FALSE,
                   Value = "value", NodeID = "names", fontSize = 12,
                   colourScale=my_color, LinkGroup="group", NodeGroup="group")

write.csv(links, file="combined.spots/Links.Cell.Populations.csv")
```
### **scRNA neuroprocesses gene signatures and spot correlation**

```{r}
#scRNA-seq genelists
load("Data/scRNA-Ruzicka_data/scRNA_GS_list.Rdata")
scRNA.list = c(scRNA_GS_list)

# scRNA.list = q$KM_G
# pdf(file = paste0(dir.name, "/", "scRNA.proc.mapping.pdf"))
# for(i in 1:length(scRNA.list)) {
#   image(matrix(log10(apply(env$metadata, 1, var)), env$preferences$dim.1stLvlSom,
#                env$preferences$dim.1stLvlSom), axes = FALSE, col = env$color.palette.heatmaps(1000),
#         main = names(scRNA.list)[i], cex.main = 1.5)
#   
#   set.genes <- unique(names(env$gene.info$names)[env$gene.info$names %in% scRNA.list[[i]] ])
#   show(length(set.genes))
#   n.map <- matrix(0, env$preferences$dim.1stLvlSom, env$preferences$dim.1stLvlSom)
#   # n.map <- rep(NA, times=env$preferences$dim.1stLvlSom*env$preferences$dim.1stLvlSom)
#   gs.nodes <- env$som.result$feature.BMU[set.genes]
#   n.map[as.numeric(names(table(gs.nodes)))] <- table(gs.nodes)
#   n.map[which(n.map == 0)] <- NA
#   n.map <- matrix(n.map, env$preferences$dim.1stLvlSom)
#   lim <- c(1, env$preferences$dim.1stLvlSom) + env$preferences$dim.1stLvlSom *
#     0.01 * c(-1, 1)
#   colr <- env$color.palette.heatmaps(1000)[(na.omit(as.vector(n.map)) -
#                                               min(n.map, na.rm = TRUE))/max(1, (max(n.map, na.rm = TRUE) - min(n.map, na.rm = TRUE))) * 999 + 1]
#   par(new=T)
#   plot(which(!is.na(n.map), arr.ind = TRUE), xlim = lim,
#        ylim = lim, pch = 16, axes = FALSE, xlab = "", ylab = "",
#        xaxs = "i", yaxs = "i", col = colr, cex = 0.5 +
#          na.omit(as.vector(n.map))/max(n.map, na.rm = TRUE) *
#          2.8)
# }
# dev.off()

scRNA.gs.def.list <- vector(mode="list", length=length(scRNA.list))

for(i in 1:length(scRNA.gs.def.list)) {
  ind <- env$gene.info$names %in% scRNA.list[[i]]
  e.genes <- unique(env$gene.info$ids[ind])
  scRNA.gs.def.list[[i]]$Genes <- e.genes
  scRNA.gs.def.list[[i]]$Type <- "scRNA"
}

names(scRNA.gs.def.list) <- names(scRNA.list)

mean.ex.all <- colMeans( env$indata.ensID.m )
sd.ex.all <- apply( env$indata.ensID.m, 2, sd )

scRNA.GSZ.scores <- t( sapply( scRNA.gs.def.list, Sample.GSZ, env$indata.ensID.m, mean.ex.all, sd.ex.all ) )

pdf(file = paste0(dir.name, "/", "scRNA.proc.boxplot.pdf"))
# par(mfrow = c(4,5))
par(mar=c(10,5,5,5))
for(i in 1:nrow(scRNA.GSZ.scores))
  boxplot(scRNA.GSZ.scores[i, ] ~ factor(env$group.labels, levels = unique(env$group.labels)), 
          col = env$groupwise.group.colors, main = rownames(scRNA.GSZ.scores)[i], border = T, 
          las=3, xlab = NULL)

dev.off()
```

```{r}
pdf(file = paste0(dir.name, "/", "scRNA.proc.correlation.pdf")) 

ind.cor <- list(ind.c = grep("Cntrl_", env$group.labels), 
                ind.s = grep("Scz_", env$group.labels), 
                ind.b = grep("BP_", env$group.labels),
                ind.m = grep("MDD_", env$group.labels))

y.list = list()

for(i in 1:length(ind.cor)){
  x <- rbind(scRNA.GSZ.scores[, ind.cor[[i]]], spot.list$spotdata[,ind.cor[[i]]])
  y <- cor(t(x))
  y <- y[rownames(spot.list$spotdata), rownames(scRNA.GSZ.scores)]
  y.list[[i]] <- y
  
  # gp <- gplots::heatmap.2(y, trace = "none", col = env$color.palette.heatmaps(1000),
  #                   Colv = "Rowv", 
  #                   scale = "none", dendrogram = "both", cexCol = 0.75, margins = c(9,5))
  }

names(y.list) <- c("Control", "SCZ", "BP", "MDD")
dev.off()
```

```{r}
comparison.list = list(
  samplesC18_35=list(C18_35=names(env$group.labels)[env$group.labels=="Cntrl_(18-35]"], 
                     S18_35=names(env$group.labels)[env$group.labels=="Scz_(18-35]"],
                     B18_35=names(env$group.labels)[env$group.labels=="BP_(18-35]"], 
                     M18_35=names(env$group.labels)[env$group.labels=="MDD_(18-35]"]),
  
  samplesC35_50=list(C35_50=names(env$group.labels)[env$group.labels=="Cntrl_(35-50]"], 
                     S35_50=names(env$group.labels)[env$group.labels=="Scz_(35-50]"],
                     B35_50=names(env$group.labels)[env$group.labels=="BP_(35-50]"],
                     M35_50=names(env$group.labels)[env$group.labels=="MDD_(35-50]"]),
  
  samplesC50_65=list(C50_65=names(env$group.labels)[env$group.labels=="Cntrl_(50-65]"], 
                     S50_65=names(env$group.labels)[env$group.labels=="Scz_(50-65]"],
                     B50_65=names(env$group.labels)[env$group.labels=="BP_(50-65]"],
                     M50_65=names(env$group.labels)[env$group.labels=="MDD_(50-65]"]),
  
  samplesC65_85=list(C65_85=names(env$group.labels)[env$group.labels=="Cntrl_(65-85]"],
                     S65_85=names(env$group.labels)[env$group.labels=="Scz_(65-85]"],
                     B65_85=names(env$group.labels)[env$group.labels=="BP_(65-85]"],
                     M65_85=names(env$group.labels)[env$group.labels=="MDD_(65-85]"]),
  samplesC85_100=list(C85_100=names(env$group.labels)[env$group.labels=="Cntrl_(85-100]"], 
                      S85_100=names(env$group.labels)[env$group.labels=="Scz_(85-100]"])
)


pdf(file = "combined.spots/spot_boxplots_by_group.pdf")
# par(mfrow = c(4,5))
par(mar=c(10,5,5,5))

for(k in 1:length(comparison.list)){
  
  samples <- unlist(comparison.list[[k]])
  gr <- env$group.labels[samples]
  cols <- env$group.colors[samples]
  
  for(i in 1:nrow(spot.list$spotdata))
    boxplot(spot.list$spotdata[i, samples] ~ factor(gr, levels = unique(gr)), 
            col = unique(cols), main = rownames(spot.list$spotdata)[i], 
            las=3, xlab = NULL)
}
dev.off()
```

## **Temporal signal in spots** 

```{r, echo=FALSE}

cntrl.group.labels = env$group.labels[grep("Cntrl", env$group.labels)]
cntrl.groupwise.group.colors <- env$groupwise.group.colors[grep("Cntrl", names(env$groupwise.group.colors))]
cntrl.spotdata <- spot.list$spotdata[, names(cntrl.group.labels)]

sd.threshold = sd(spot.list$spotdata)
pdf("combined.spots/Control.Spot.Boxplots.pdf")
layout(matrix(1:8, 4, 2, byrow = TRUE), widths = c(1, 4))
ylim <- range(unlist(by(t(cntrl.spotdata), cntrl.group.labels, 
                        range)[unique(cntrl.group.labels)]))
for (i in 1:length(spot.list$spots)) {
  par(mar = c(2, 5, 0.5, 1))
  image(matrix(spot.list$spots[[i]]$mask, env$preferences$dim.1stLvlSom, 
               env$preferences$dim.1stLvlSom), axes = FALSE, col = "darkgreen")
  axis(2, 0.95, names(spot.list$spots[i]), las = 2, tick = FALSE, 
       cex.axis = 1)
  box()
  par(mar = c(2, 3, 0.5, 1))
  boxplot(tapply(cntrl.spotdata[i, ], cntrl.group.labels, 
                 c)[unique(cntrl.group.labels)], col = cntrl.groupwise.group.colors, 
          las = 1, ylim = ylim)
  abline(h = sd.threshold * c(-1, 0, 1), lty = 2, col = "gray")
  label.string <- rep(".", length(unique(cntrl.group.labels)))
  label.string[which(tapply(cntrl.spotdata[i, ], cntrl.group.labels, 
                            mean)[unique(cntrl.group.labels)] > sd.threshold)] = "+"
  label.string[which(tapply(cntrl.spotdata[i, ], cntrl.group.labels, 
                            mean)[unique(cntrl.group.labels)] < -sd.threshold)] = "-"
  text(1:length(unique(cntrl.group.labels)), par("usr")[4] - 
         (par("usr")[4] - par("usr")[3]) * 0.05, label.string, 
       col = cntrl.groupwise.group.colors, cex = 3)
}
dev.off()
```

**Function for data preparation**

```{r, include=FALSE}
prepare.data <- function(what="Cntrl", where="Data/pheno", group_labels = env$group.labels ){ #q redundant with PCA chunk, define earlier, use once
  
  ind <- grepl(pattern = what, group_labels)
  ind <- which(ind)
  mat <- spot.list$spotdata[, ind]
  batch <- env$combat.batch[ind]
  group.labels <- group_labels[ind]
  names(batch) <- names(group.labels)
  
  pheno_vars <- load_pheno(where=where,batch)
  age <- pheno_vars[[1]]
  sex <- pheno_vars[[2]]
  ph <- pheno_vars[[1]]
  pmi <- pheno_vars[[1]]
  
  pmi[is.na(pmi)] <- mean(pmi, na.rm = T) #q these 2 lines were not put inside this function, added, remove if not needed
  ph[is.na(ph)] <- mean(ph, na.rm = T)
  
  col.pal <- brewer.pal(length(unique(batch)), name = "Dark2")
  delta <- min(batch)-1
  cols <- col.pal[batch-delta]
  
  what.list = list(mat=mat,age=age, sex=sex, ph=ph, pmi=pmi)
}
```

# **Prepare the Data for downstream analysis**

```{r, include=FALSE}
# prepare data for analysis
cntrl.list <- prepare.data(what="Cntrl", where="Data/pheno")
scz.list <- prepare.data(what="Scz", where="Data/pheno")
bp.list <- prepare.data(what="BP", where="Data/pheno")
mdd.list <- prepare.data(what="MDD", where="Data/pheno")

ControlTimes <- cntrl.list$age
ControlData <- cntrl.list$mat

PerturbedTimes.scz <- scz.list$age
PerturbedData.scz <- scz.list$mat

PerturbedTimes.bp <- bp.list$age
PerturbedData.bp <- bp.list$mat

PerturbedTimes.mdd <- mdd.list$age
PerturbedData.mdd <- mdd.list$mat

o <- order(ControlTimes, decreasing = FALSE)
ControlData <- ControlData[, o]
ControlTimes <- ControlTimes[o]

o1 <- order(PerturbedTimes.scz, decreasing = FALSE)
PerturbedData.scz <- PerturbedData.scz[, o1]
PerturbedTimes.scz <- PerturbedTimes.scz[o1]
TestTimes.scz <- seq(min(c(min(ControlTimes), min(PerturbedTimes.scz))), max(c(max(ControlTimes), max(PerturbedTimes.scz))), length=100)

o2 <- order(PerturbedTimes.bp, decreasing = FALSE)
PerturbedData.bp <- PerturbedData.bp[, o2]
PerturbedTimes.bp <- PerturbedTimes.bp[o2]
TestTimes.bp <- seq(min(c(min(ControlTimes), min(PerturbedTimes.bp))), max(c(max(ControlTimes), max(PerturbedTimes.bp))), length=100)

o3 <- order(PerturbedTimes.mdd, decreasing = FALSE)
PerturbedData.mdd <- PerturbedData.mdd[, o3]
PerturbedTimes.mdd <- PerturbedTimes.mdd[o3]
TestTimes.mdd <- seq(min(c(min(ControlTimes), min(PerturbedTimes.mdd))), max(c(max(ControlTimes), max(PerturbedTimes.mdd))), length=100)
```

# **Linear regression pairwise age-disease comparisons**

```{r}
p.vals.lm <- vector(mode="numeric", length=nrow(ControlData))
p.vals.mod <- vector(mode="numeric", length=nrow(ControlData))
fit.list <- list()
for(i in 1:length(p.vals.lm)){
  fit <- lm(ControlData[i,]~ControlTimes)
  p.vals.lm[i] <- summary(fit)$coefficients[2,4]
  p.vals.mod[i] <- pf(summary(fit)$fstatistic[1],
                      summary(fit)$fstatistic[2],
                      summary(fit)$fstatistic[3],
                      lower.tail = FALSE)
  fit.list[[i]] <- fit
}

names(fit.list) <- rownames(ControlData)
p.vals.lm.adj <- round(p.adjust(p.vals.lm, method = "bonferroni"), digits = 6)
names(p.vals.lm.adj) <- rownames(ControlData)
```

```{r}
# SCZ
comp.data.scz <- cbind(ControlData, PerturbedData.scz)

mod.scz <- data.frame(status=factor(c(rep("control", times=ncol(ControlData)), 
                                      rep("scz", times=ncol(PerturbedData.scz))), 
                                    levels = c("control", "scz") ), 
                      age=c(ControlTimes, PerturbedTimes.scz))

ind.1 <- mod.scz$age>=17 

comp.data.scz <- comp.data.scz[, ind.1] 
mod.scz <- mod.scz[ind.1, ]

fit.list.scz <- list()
p.vals.scz.lm <- vector(mode="numeric", length=nrow(comp.data.scz))
names(p.vals.scz.lm) <- rownames(comp.data.scz)

for(i in 1:nrow(comp.data.scz)) {
  fit.scz <- lm(comp.data.scz[i,]~mod.scz$status+mod.scz$age)
  p.vals.scz.lm[i] <- summary(fit.scz)$coefficients[2,4]
  fit.list.scz[[i]] <- fit.scz
}

p.vals.scz.lm.adj <- round(p.adjust(p.vals.scz.lm, method = "bonferroni"), digits = 4)
names(fit.list.scz) <- names(p.vals.scz.lm)
```

```{r}
# BP

comp.data.bp <- cbind(ControlData, PerturbedData.bp)

mod.bp <- data.frame(status=factor(c(rep("control", times=ncol(ControlData)), 
                                     rep("bp", times=ncol(PerturbedData.bp))), 
                                   levels = c("control", "bp") ), 
                     age=c(ControlTimes, PerturbedTimes.bp))

ind.1 <- mod.bp$age>=17

comp.data.bp <- comp.data.bp[, ind.1] 
mod.bp <- mod.bp[ind.1, ]

fit.list.bp <- list()
p.vals.bp.lm <- vector(mode="numeric", length=nrow(comp.data.bp))
names(p.vals.bp.lm) <- rownames(comp.data.bp)

for(i in 1:nrow(comp.data.bp)) {
  fit.bp <- lm(comp.data.bp[i,]~mod.bp$status+mod.bp$age)
  p.vals.bp.lm[i] <- summary(fit.bp)$coefficients[2,4]
  fit.list.bp[[i]] <- fit.bp
}

p.vals.bp.lm.adj <- round(p.adjust(p.vals.bp.lm, method = "bonferroni"), digits = 4)
names(fit.list.bp) <- names(p.vals.bp.lm)
```

```{r}
# MDD

comp.data.mdd <- cbind(ControlData, PerturbedData.mdd)

mod.mdd <- data.frame(status=factor(c(rep("control", times=ncol(ControlData)), 
                                      rep("mdd", times=ncol(PerturbedData.mdd))), 
                                    levels = c("control", "mdd") ), 
                      age=c(ControlTimes, PerturbedTimes.mdd))

ind.1 <- mod.mdd$age>=17

comp.data.mdd <- comp.data.mdd[, ind.1] 
mod.mdd <- mod.mdd[ind.1, ]

fit.list.mdd <- list()
p.vals.mdd.lm <- vector(mode="numeric", length=nrow(comp.data.mdd))
names(p.vals.mdd.lm) <- rownames(comp.data.mdd)

for(i in 1:nrow(comp.data.mdd)) {
  fit.mdd <- lm(comp.data.mdd[i,]~mod.mdd$status+mod.mdd$age)
  p.vals.mdd.lm[i] <- summary(fit.mdd)$coefficients[2,4]
  fit.list.mdd[[i]] <- fit.mdd
}

p.vals.mdd.lm.adj <- round(p.adjust(p.vals.mdd.lm, method = "bonferroni"), digits = 4)
names(fit.list.mdd) <- names(p.vals.mdd.lm)

lm.spots <- list(SCZ=names(p.vals.scz.lm.adj)[p.vals.scz.lm.adj<=0.05], 
                 BP=names(p.vals.bp.lm.adj)[p.vals.bp.lm.adj<=0.05],
                 MDD=names(p.vals.mdd.lm.adj)[p.vals.mdd.lm.adj<=0.05])


venn::venn(lm.spots, zcolor = "style")
```

# **Gaussian process regression**

<https://cran.r-project.org/web/packages/GauPro/vignettes/GauPro.html>

<https://www.r-bloggers.com/2012/04/gaussian-process-regression-with-r/>

<https://mc-stan.org/docs/2_19/stan-users-guide/gaussian-process-regression.html>

<https://bioinformatics-training.github.io/intro-machine-learning-2017/logistic-regression.html#gaussian-process-regression>

### **DE_time differential perturbation**

DEtime is an R package for two-sample time series analysis using Gaussian process methods. This package implements the Gaussian regression framework for perturbation time point inferrence in a two sample case. The paper describing this package is available at DOI: <https://doi.org/10.1093/bioinformatics/btw329> and arXiv: <http://arxiv.org/abs/1602.01743>.

```{r echo=FALSE}
# In case package archive installation fails

# r.files <- dir("Tools/gptk/R/", full.names = T)
# sapply(r.files, source)
# 
# r.files.1 <- dir("Tools/DEtime/R/", full.names = T)
# sapply(r.files.1, source) 
```

```{r echo=FALSE}
# SCZ
res_rank.scz <- DEtime_rank(ControlTimes, ControlData, PerturbedTimes.scz, PerturbedData.scz, gene_ID = rownames(ControlData))
res.scz <- DEtime_infer(ControlTimes = ControlTimes, ControlData = ControlData, PerturbedTimes = PerturbedTimes.scz, PerturbedData = PerturbedData.scz, TestTimes = TestTimes.scz, gene_ID = rownames(ControlData))

res.df.scz <- data.frame(round(res_rank.scz$Loglikelihood_ratio, digits = 2), 
                         as.numeric(res.scz$result$MAP), 
                         as.numeric(res.scz$result$mean), 
                         as.numeric(res.scz$result$median), 
                         as.numeric(res.scz$result$ptl5), 
                         as.numeric(res.scz$result$ptl95) )

colnames(res.df.scz) <- c("log-likelihood ratio", "MAP", "mean", "median", "ptl9", "ptl95")
rownames(res.df.scz) <- res_rank.scz$gene_ID
res.df.scz <- res.df.scz[res.df.scz$`log-likelihood ratio` >= 1, ] 
# Plotting GPR
# plot_DEtime(res.scz)
## END SCZ
```

```{r echo=FALSE}
# BP
res_rank.bp <- DEtime_rank(ControlTimes, ControlData, PerturbedTimes.bp, PerturbedData.bp, gene_ID = rownames(ControlData))
res.bp <- DEtime_infer(ControlTimes = ControlTimes, ControlData = ControlData, PerturbedTimes = PerturbedTimes.bp, PerturbedData = PerturbedData.bp, TestTimes = TestTimes.bp, gene_ID = rownames(ControlData))

res.df.bp <- data.frame(round(res_rank.bp$Loglikelihood_ratio, digits = 2), 
                        as.numeric(res.bp$result$MAP), 
                        as.numeric(res.bp$result$mean), 
                        as.numeric(res.bp$result$median), 
                        as.numeric(res.bp$result$ptl5), 
                        as.numeric(res.bp$result$ptl95) )

colnames(res.df.bp) <- c("log-likelihood ratio", "MAP", "mean", "median", "ptl9", "ptl95")
rownames(res.df.bp) <- res_rank.bp$gene_ID
res.df.bp <- res.df.bp[res.df.bp$`log-likelihood ratio` >= 1, ] 
# Plotting GPR
# plot_DEtime(res.bp)
## END BP
```

```{r echo=FALSE}
# MDD
res_rank.mdd <- DEtime_rank(ControlTimes, ControlData, PerturbedTimes.mdd, PerturbedData.mdd, gene_ID = rownames(ControlData))
res.mdd <- DEtime_infer(ControlTimes = ControlTimes, ControlData = ControlData, PerturbedTimes = PerturbedTimes.mdd, PerturbedData = PerturbedData.mdd, TestTimes = TestTimes.mdd, gene_ID = rownames(ControlData))

ind <- res_rank.mdd$Loglikelihood_ratio > 1
res.df.mdd <- data.frame(round(res_rank.mdd$Loglikelihood_ratio, digits = 2), 
                         as.numeric(res.mdd$result$MAP), 
                         as.numeric(res.mdd$result$mean), 
                         as.numeric(res.mdd$result$median), 
                         as.numeric(res.mdd$result$ptl5), 
                         as.numeric(res.mdd$result$ptl95) )

colnames(res.df.mdd) <- c("log-likelihood ratio", "MAP", "mean", "median", "ptl9", "ptl95")
rownames(res.df.mdd) <- res_rank.mdd$gene_ID
res.df.mdd <- res.df.mdd[res.df.mdd$`log-likelihood ratio` >= 1, ] 

# Plotting GPR
# plot_DEtime(res.mdd)
## END MDD
```

```{r echo=FALSE}
DE.spots <- list(SCZ=rownames(res.df.scz), 
                 BP=rownames(res.df.bp),
                 MDD=rownames(res.df.mdd) )

venn::venn(DE.spots, zcolor = "style")

early.DE.spots <- list(SCZ=rownames(res.df.scz)[res.df.scz$MAP<8],
                       BP=rownames(res.df.bp)[res.df.bp$MAP<8],
                       MDD=rownames(res.df.mdd)[res.df.mdd$MAP<8])


venn::venn(early.DE.spots, zcolor = "style")

pub.DE.spots <- list(SCZ=rownames(res.df.scz)[res.df.scz$MAP>8 & res.df.scz$MAP<20],
                       BP=rownames(res.df.bp)[res.df.bp$MAP>8 & res.df.bp$MAP<20],
                       MDD=rownames(res.df.mdd)[res.df.mdd$MAP>8 & res.df.mdd$MAP<20])

venn::venn(pub.DE.spots, zcolor = "style")

adol.DE.spots <- list(SCZ=rownames(res.df.scz)[res.df.scz$MAP>20 & res.df.scz$MAP<50],
                       BP=rownames(res.df.bp)[res.df.bp$MAP>20 & res.df.bp$MAP<50],
                       MDD=rownames(res.df.mdd)[res.df.mdd$MAP>20 & res.df.mdd$MAP<50])

venn::venn(adol.DE.spots, zcolor = "style")

late.DE.spots <- list(SCZ=rownames(res.df.scz)[res.df.scz$MAP>50],
                       BP=rownames(res.df.bp)[res.df.bp$MAP>50],
                       MDD=rownames(res.df.mdd)[res.df.mdd$MAP>50])

venn::venn(late.DE.spots, zcolor = "style")
```



# **QTL enrichment analysis**

```{r echo=FALSE}
load("Data/BM.table.Rdata") # gene data annotation table

gtex.files <- dir("Data/GTEX_unzipped", full.names = TRUE) 
gtex.files <- gtex.files[-c(21, 22)]
gtex.files.names <- sapply(gtex.files, function(x) strsplit(x = x,split = "/", fixed = T)[[1]][3])
gtex.files.names <- sapply(gtex.files.names, function(x) 
  strsplit(x = x,split = ".", fixed = T)[[1]][1])
names(gtex.files.names) <- NULL

spot.names <- rownames(ControlData)

qtl.spot.list <- list()
p.list <- vector(mode = "list", length = length(spot.names))
names(p.list) <- spot.names

for(spot in spot.names){
  spot.probes <- spot.list$spots[[spot]]$genes
  
  spot.genes <- unique(BM.table$hgnc_symbol[BM.table$affy_hg_u133_plus_2 %in% spot.probes])
  spot.genes <- spot.genes[spot.genes!=""]
  
  p <- vector(mode = "numeric", length = length(gtex.files))
  qtl.gene.list <- list()
  
  for(i in 1:length(gtex.files)) {
    qtl <- read.table(gtex.files[i], sep = "\t", header = TRUE, row.names = 1)
    
    # qtl genes
    qtl.genes <- unique(qtl$gene_name[qtl$qval <= 0.05])
    qtl.genes.backgroud <- unique(env$gene.info$names[env$gene.info$names %in% qtl.genes])
    
    # non qtl genes
    nonqtl.genes.backgroud <- setdiff(unique(BM.table$hgnc_symbol),qtl.genes.backgroud)
    nonqtl.genes.backgroud <- nonqtl.genes.backgroud[nonqtl.genes.backgroud!=""]
    
    qtl.spot.genes <- intersect(spot.genes, qtl.genes.backgroud)
    nonqtl.spot.genes <- setdiff(spot.genes, qtl.spot.genes)
    
    qtl.mat <- matrix(c(length(qtl.spot.genes), length(nonqtl.spot.genes),
                        length(qtl.genes.backgroud)-length(qtl.spot.genes),
                        length(nonqtl.genes.backgroud)-length(nonqtl.spot.genes)),
                      2,2)
   
    p[i] <- fisher.test(qtl.mat, alternative = "greater")$p.value
    
    qtl.gene.list[[i]] <- qtl.spot.genes
    
  }
  
  p.adj <- round(p.adjust(p, method = "bonferroni"), digits = 4)
  names(p) <- names(p.adj) <- names(qtl.gene.list) <- gtex.files.names
  p.list[[spot]] <- list(p = p, p.adj = p.adj)
  qtl.spot.list[[spot]] <- qtl.gene.list
  
}

tissues.all <- sapply(p.list, function(x) names(x$p.adj)[x$p.adj<=0.05])
tissue.counts <- sapply(tissues.all, function(x) length(x) )
tissues.all <- tissues.all[tissue.counts!=0]

scz.spots <- DE.spots$SCZ
bp.spots <- DE.spots$BP
mdd.spots <- DE.spots$MDD

scz.qtl.genes <- unique(unlist(qtl.spot.list[names(qtl.spot.list) %in% scz.spots]))
bp.qtl.genes <- unique(unlist(qtl.spot.list[names(qtl.spot.list) %in% bp.spots]))
mdd.qtl.genes <- unique(unlist(qtl.spot.list[names(qtl.spot.list) %in% mdd.spots]))

setEnrichrSite("Enrichr")
websiteLive <- TRUE
dbs <- listEnrichrDbs()

query.dbs <- dbs$libraryName[c(159, 144, 155)] #dbs - DisGENET, GWAS catalog, UK Biobank GWAS

if (websiteLive) {
  enriched.scz <- enrichr(scz.qtl.genes, query.dbs)
  enriched.bp <- enrichr(bp.qtl.genes, query.dbs)
  enriched.mdd <- enrichr(mdd.qtl.genes, query.dbs)
}

yy.scz <- lapply(enriched.scz, function(x) {
  ind <- grep("schizophre", x$Term, ignore.case = TRUE)
  y <- x[ind, ]
  y <- y[y$Adjusted.P.value < 0.1, ]
  return(y)
})

ind <- sapply(yy.scz, function(x) nrow(x) )>0

yy.scz <- yy.scz[ind]

yy.scz.df <- do.call(rbind, yy.scz)

yy.bp <- lapply(enriched.bp, function(x) {
  ind <- grep("bipolar", x$Term, ignore.case = TRUE)
  y <- x[ind, ]
  y <- y[y$Adjusted.P.value < 0.1, ]
  return(y)
})

ind <- sapply(yy.bp, function(x) nrow(x) )>0

yy.bp <- yy.bp[ind]

yy.bp.df <- do.call(rbind, yy.bp)

yy.mdd <- lapply(enriched.mdd, function(x) {
  ind <- grep("depres", x$Term, ignore.case = TRUE)
  y <- x[ind, ]
  y <- y[y$Adjusted.P.value < 0.1, ]
  return(y)
})

ind <- sapply(yy.mdd, function(x) nrow(x) )>0

yy.mdd <- yy.mdd[ind]

yy.mdd.df <- do.call(rbind, yy.mdd)

dis.geneetic.association.df <- rbind(yy.scz.df, yy.bp.df, yy.mdd.df)
write.csv(dis.geneetic.association.df, file="dis.geneetic.association.df.csv")

tissues <- tissues.all[names(tissues.all) %in% unique(c(scz.spots, bp.spots, mdd.spots))]

blood <- unlist(lapply(tissues, function(x) {sum(grepl(pattern = "blood", x = x, 
                                                       ignore.case = T))}))
brain <- unlist(lapply(tissues, function(x) {sum(grepl(pattern = "cortex", x = x, 
                                                       ignore.case = T))}))


```

-   Sankeyplot spot-QTL relation

```{r, echo=FALSE}
nodes <- c(sort(unique(unlist(tissues))), names(tissues))
nodes.rank <- seq(0,length(nodes), by=1)
names(nodes.rank) <- nodes
links <- NULL
for(i in 1:length(tissues)) {
  tmp <- tissues[[i]]
  if(length(tmp)==0)
    next
  for(j in 1:length(tmp)){
    if(j != 0){
      spot.tmp <- names(tissues)[i]
      tissue.tmp <- tmp[j]
      genes.tmp <- length(qtl.spot.list[[spot.tmp]][[tissue.tmp]])
      links <- rbind(links, c(nodes.rank[tissue.tmp], nodes.rank[spot.tmp], genes.tmp))
    }
  }
}
colnames(links) = c("source", "target", "value")
links <- data.frame(links)
nodes.df <- data.frame(name = c(sort(unique(unlist(tissues))), names(tissues)))

p = sankeyNetwork(Links = links, Nodes = nodes.df,
                  Source = "source", Target = "target",
                  Value = "value", NodeID = "name",
                  fontSize = 12, nodeWidth = 20)
```

